%% =====================================================================
%% From Passive Members to Self-Aware Steel Structures
%% Target journal: Automation in Construction (Elsevier, ISSN 0926-5805)
%% Class: elsarticle (numbered references)
%% Compile: latexmk -pdf -interaction=nonstopmode main.tex
%% =====================================================================
\documentclass[preprint,12pt]{elsarticle}

\usepackage{amssymb,amsmath}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,shapes.geometric,shapes.misc,calc,fit,backgrounds,decorations.pathmorphing}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{groupplots}

%% Elsevier-friendly colour palette (used by the TikZ/pgfplots figures)
\definecolor{elsB1}{HTML}{08306B}
\definecolor{elsB2}{HTML}{2171B5}
\definecolor{elsB3}{HTML}{6BAED6}
\definecolor{elsB4}{HTML}{C6DBEF}
\definecolor{elsC1}{HTML}{66C2A5}
\definecolor{elsC2}{HTML}{FC8D62}

%% For double line spacing during review, use the [review] class option.
%% \usepackage{lineno}\linenumbers

\journal{Automation in Construction}

\begin{document}

\begin{frontmatter}

\title{From Passive Members to Self-Aware Steel Structures: A Digital-Twin and
Machine-Learning Framework for Structural Cognition and Autonomous Health Monitoring}

\author[inst1]{[First Author]\corref{cor1}}
\ead{first.author@institution.edu}

\author[inst1]{[Second Author]}

\cortext[cor1]{Corresponding author.}

\affiliation[inst1]{organization={[Department]},
            addressline={[Street address]},
            city={[City]},
            postcode={[Postcode]},
            state={[State]},
            country={[Country]}}

\begin{abstract}
Steel structures are conventionally treated as passive load-resisting members
whose integrity is verified by periodic inspection. Structural health monitoring
(SHM) added continuous measurement, yet many deployments still stop at data
acquisition, leaving interpretation to offline analysis. This paper proposes a
conceptual framework and reference architecture for the Self-Aware Steel
Structure, which perceives, processes, interprets and autonomously reports its
own mechanical state through a permanent link to a continuously updated digital
twin. We introduce the paradigm of Structural Cognition and an operational
six-level scale (L0--L5) of structural self-awareness mapped onto the Rytter
damage hierarchy. A five-layer architecture couples distributed multi-sensor
acquisition, a finite-element digital twin, a machine-learning cognition layer
(unsupervised novelty detection with autoencoders, supervised classification, and
LSTM temporal modelling) and a governed decision layer. A braced
square-hollow-section frame under rotating-unbalance excitation is specified as
the validation testbed, advancing cognitive digital twins and autonomous
predictive maintenance of steel infrastructure.
\end{abstract}

\begin{graphicalabstract}
\centering
\resizebox{0.96\linewidth}{!}{\input{graphical-abstract-body}}
\end{graphicalabstract}

\begin{highlights}
\item Self-Aware Steel Structure paradigm shifts members from passive to cognitive.
\item Structural Cognition lets a structure interpret and report its own state.
\item Five-layer architecture couples sensing, FEM digital twin and learning.
\item Supervised and unsupervised AI target stiffness loss and bolt-loosening detection.
\item SHS braced-frame testbed enables autonomous predictive maintenance of steel.
\end{highlights}

\begin{keyword}
Self-aware structures \sep Structural cognition \sep Digital twin \sep
Structural health monitoring \sep Machine learning \sep
Autonomous damage detection \sep Steel structures
\end{keyword}

\end{frontmatter}

%% ===================== Main text =====================
\input{secoes/01-introduction}
\input{tabelas/tab1-pillars}
\input{figs/fig-architecture}

\input{secoes/02-background}
\input{figs/fig-frf}
\input{tabelas/tab-positioning}

\input{secoes/03-paradigm}
\input{tabelas/tab2-conv-vs-aware}
\input{figs/fig-levels}
\input{tabelas/tab3-levels}

\input{secoes/04-architecture}
\input{tabelas/tab4-sensors}

\input{secoes/05-methodology}
\input{figs/fig-testbed}
\input{tabelas/tab6-params}

\input{secoes/06-ml-strategy}
\input{figs/fig-mlpipeline}
\input{tabelas/tab5-mlmethods}

\input{secoes/07-discussion}

\input{secoes/08-conclusions}

%% ===================== Declarations =====================
\section*{CRediT authorship contribution statement}
\textbf{[First Author]:} Conceptualization, Methodology, Writing -- original draft.
\textbf{[Second Author]:} Conceptualization, Supervision, Writing -- review \& editing.

\section*{Declaration of competing interest}
The authors declare that they have no known competing financial interests or
personal relationships that could have appeared to influence the work reported in
this paper.

%% ===================== Bibliography =====================
\bibliographystyle{elsarticle-num}
\bibliography{references}

\end{document}
